Variable Density Tesseroids: Gravity fields calculation in spherical coordinates using variable densities in depth

### 2017-10-18 manuscript/manuscript.tex
Changes in manuscript: modified determination of D

In order to test the new variable density tesseroid code and estimate the error of the approximation, we have compared the numerical model with the analytical solutions for a spherical shell with variable density for three special cases: a linear, an exponential and a discontinuous density function.
The first one constitutes the simplest variation, while the second one represents the fastest varying function. The later one tries to emulate a contact between two different media (e.g. crust and mantle interface).

This error estimations are also useful in the determination of the minimum value for the distance-size ratio $D$ such the difference between the numerical model and the analytical solutions are below an acceptable threshold.

@@

We computed these gravity fields  on every point of the four grids for different values of $D$, ranging from 0.5 to 10 with a step of 0.5.
Then we calculated the maximum difference between these results and the value of the analytical solution at the same height.

@@

@@

\subsection{Linear Density}

In first place, we proposed a spherical shell with linear density in such way that it has a density of 2670~kg/m$^3$ on the surface of the Earth, and 3300~kg/m$^3$ at 35 km of depth:

\begin{equation}
    \rho(r') = ar' + c,
    \label{eq:density-linear}
\end{equation}
\noindent with 
\begin{equation}
    a = -\frac{3300\text{kg/m$^3$} - 2670\text{kg/m$^3$}}{35000\text{m}},
\end{equation}
\begin{equation}
    c = \frac{3300\text{kg/m$^3$} - 
        2670\text{kg/m$^3$}}{35000\text{m}} R + 
        2670\text{kg/m$^3$},
\end{equation}

@@

\begin{equation}
    \rho(r') = A e^{-(r' - R)/b}
\label{eq:density-exp}
\end{equation}

\noindent where $A$ = 1kg/m$^3$ and $b$ = 1000km.
We have performed this test also with a thin and a thick shell of 1 km and 35 km, respectively, whose results can be seen in Figures \ref{fig:D-exp-thin} and \ref{fig:D-exp-thick}.
Once again, the results for $g_{xx}$ and $g_{yy}$ are very similar to the ones obtained for $g_{zz}$, so we exclude them, although, they can be found in the repository.

In the same way we analysed the linear density results, we can summarise that the thin spherical shell presents the more conservative values of the distance-size ratio: $D=1$ for the gravity potential, $D=2$ for its gradient components and $D=8$ for the Marussi tensor components.

@@

Because the constant $b$ controls the variation of the density function, we wanted to test how does the accuracy of the method behaves at different values of $b$.
In order to do this we proposed a set of density functions, each one with a different value of $b$ and normalised in such way that they reaches a density of 2900 kg/m$^{3}$ at the bottom of the spherical shell.
We computed the difference between the Marussi tensor component $g_{zz}$ generated by the numerical model on the big grid of 30$^\circ\times$30$^\circ$ (see Table \ref{tab:grids}) and the respective analytical solution.
These differences were calculated for each proposed density function and for the same set of values of $D$ we have been using on the previous tests, obtaining one curve for each value of $b$.

This tests were performed using both the thin and the thick spherical shell.
On Figures \ref{fig:D-exp-power-thin}a and \ref{fig:D-exp-power-thick}a we can see the density functions used to perform these tests for each spherical shell model, respectively, and their results can be seen in \ref{fig:D-exp-power-thin}b and \ref{fig:D-exp-power-thick}b.
Taking into account that lower values of $b$ implies a more varying function, we can see that for extremely varying functions the numerical model does not perform well: for $b$=100m the difference is almost 100\% in the thin shell case, and the same happens with $b$=10km in the thick shell case.

On both cases, the model performs under the acceptable threshold for values of $b$ greater than the thickness of the shell, in other words, when the absolute value of the exponent in equation \ref{eq:density-exp} is lower than one.
Although this sets a limitation to the model, the maximum acceptable variation is very far from real density distributions inside the Earth.
For example, if we take into account a tesseroid with 35km of thickness, a density of 2670kg/m$^3$ on its top surface and $b=35$km, then the density at its bottom will be approximately of 7000kg/m$^3$, very far from the values we can find at such depths.

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{figures/Dexp-power-differences-thin.pdf}
\caption{
    Results of the accuracy tests for different exponential density functions in case of a thin spherical shell (1km of thickness).
    (a) Density functions used in the tests, each one with a different value of $b$ and normalised in such way that they reaches a density of 2900kg/m$^{3}$ at the bottom of the shell.
    (b) Differences between the Marussi tensor component $g_{zz}$ generated by the numerical model and the analytical solution for each density function. Each computation was performed on the big grid described in Table \ref{tab:grids}, for different values of the distance-size ratio $D$. If the difference is less than 0.1\%, we consider that the model has achieved an acceptable accuracy for the corresponding value of $D$.
}
\label{fig:D-exp-power-thin}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{figures/Dexp-power-differences-thick.pdf}
\caption{
    Results of the accuracy tests for different exponential density functions in case of a thick spherical shell (35km of thickness).
    (a) Density functions used in the tests, each one with a different value of $b$ and normalised in such way that they reaches a density of 2900kg/m$^{3}$ at the bottom of the shell.
    (b) Differences between the Marussi tensor component $g_{zz}$ generated by the numerical model and the analytical solution for each density function. Each computation was performed on the big grid described in Table \ref{tab:grids}, for different values of the distance-size ratio $D$. If the difference is less than 0.1\%, we consider that the model has achieved an acceptable accuracy for the corresponding value of $D$.
}
\label{fig:D-exp-power-thick}
\end{figure}

@@

\noindent where $\rho_1$ and $\rho_2$ are constant densities, $R_1$ and $R_2$ are the inner and outer radii of the shell, respectively, and $R_c$ is the the radius on which the 
density changes.

The analytical solution for this kind of density is the linear combination of two homogeneous shells with densities $\rho_1$ and $\rho_2$:

@@

Firstly, lets assume a symmetric density function, i.e. $R_c=(R_1 + R_2)/2$ is the centre radius of the shell, and perform the test for a thin shell of 1km of thickness.
The results can be seen in Figure \ref{fig:D-discont-sym}.
Again, we obtain good accuracy for values of $D$ equal to 1, 2 and 8 for the computation of the potential, the gradient components and the Marussi tensor components, respectively.

@@

From equation \ref{eq:glq-var-dens} we have seen that the information of the density function is summarised as the values its assumes on the nodes of the GLQ.
We can attribute the low performance of the code to this simplification.

On the other hand, we can explain the good performance in the symmetrical density case due to the fact that the GLQ nodes are always symmetrical with respect to the centre of the $[a,b]$ integration interval (see equation \ref{eq:glq-scaled}).

In the light of these results we do not recommend using discontinuous densities functions. Instead, divide the tesseroid into several ones with specific homogeneous density or density function for each one.

>>>>

In order to determine the minimum value for the distance-size ratio $D$ such the numerical approximation achieves an acceptable accuracy we have compared the numerical model with the analytical solutions for a spherical shell with variable density for three special cases: a linear, an exponential and a discontinuous density function.
The first one constitutes the simplest variation, while the second one represents the fastest varying function. The later one tries to emulate a contact between two different media (e.g. crust and mantle interface).

@@

We computed these gravity fields  on every point of the four grids for different values of $D$, ranging from 0.5 to 10 with a step of 0.5.
Then we calculated the maximum difference between these results and the value that the analytical solution assumes at the same height.

@@

Moreover, we did not apply the density-based discretization algorithm in these tests in order to isolate the behaviour of the adaptive discretization algorithm.

@@

\subsection{Linear Density}
Firstly,  we proposed a spherical shell with linear density in such way that it assumes 2670~kg/m$^3$ on the surface of the Earth and 3300~kg/m$^3$ at the inner radius:

\begin{equation}
    \rho(r') = ar' + c,
    \label{eq:density-linear}
\end{equation}
\noindent with 
\begin{equation}
    a = -\frac{3300\text{kg/m$^3$} - 2670\text{kg/m$^3$}}{R_2 - R_1},
\end{equation}
\begin{equation}
    c = \frac{3300\text{kg/m$^3$} - 
        2670\text{kg/m$^3$}}{R_2 - R_1} R + 
        2670\text{kg/m$^3$},
\end{equation}

@@

\begin{equation}
    \rho(r') = A \left( e^{-(r' - R)/b} - 1\right) + 2670 \text{kg/m}^3
\label{eq:density-exp}
\end{equation}
\noindent where
\begin{equation}
    A =
    (3300 \text{kg/m}^3 - 2670 \text{kg/m}^3)
    \left( e^{( R_2 - R_1 )/b} - 1 \right)^{-1},
\end{equation}

\noindent and $b$ = 1000km. 
In this way, the shell has a density of 2670 kg/m$^3$ and 3300 kg/m$^3$ on its outer and inner radii, respectively.
The analytical solution of the gravity potential is easily obtained as the sum the solution for a pure exponential function and the one for an homogeneous spherical shell.

We have performed this test also with a thin and a thick shell of 1 km and 35 km, respectively, without applying the density-based discretization algorithm. Its results can be seen in Figures \ref{fig:D-exp-thin} and \ref{fig:D-exp-thick}.
Once again, the results for $g_{xx}$ and $g_{yy}$ are very similar to the ones obtained for $g_{zz}$, so we exclude them, although, they can be found in the repository.

In the same way we analysed the linear density results, we can summarise that the thin spherical shell presents the more conservative values of the distance-size ratio: $D=1$for the gravity potential, $D=2$ for its gradient components and $D=8$ for the Marussi tensor components.

@@

Because the constant $b$ controls the variation of the density function, we wanted to test how does the accuracy of the method behaves at different values of $b$ when the density-based discretization algorithm is not applied.
In order to do this we proposed a set of density functions, each one with a different value of $b$.
We computed the difference between the Marussi tensor component $g_{zz}$ generated by the numerical model on the big grid of 30$^\circ\times$30$^\circ$ (see Table \ref{tab:grids}) and the respective analytical solution.
These differences were calculated for each proposed density function and for the same set of values of $D$ we have been using on the previous tests, obtaining one curve for each value of $b$.

On Figure \ref{fig:D-exp-power-thin}a we can see the density functions used to perform these tests and its corresponding results on \ref{fig:D-exp-power-thin}b .
Taking into account that lower values of $b$ implies a more varying function, we can see that for high varying functions the numerical model does not perform well: for $b$=100m the difference is almost 100\%.

We can conclude that even when the density-based discretization algorithm is not applied, the numerical approximation performs under the acceptable threshold for values of $b$ greater than the thickness of the shell, i.e. for low density variations.
On the other hand, a density-based discretization algorithm is needed in order to guarantee the accuracy of the approximation in case of a high varying density function.

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{figures/Dexp-power-differences-thin.pdf}
\caption{
    Results of the accuracy tests for different exponential density functions in case of a spherical shell of 1km of thickness.
    (a) Density functions used in the tests, each one with a different value of $b$ and normalised in such way that they reaches a density of 2900kg/m$^{3}$ at the bottom of the shell.
    (b) Differences between the Marussi tensor component $g_{zz}$ generated by the numerical model and the analytical solution for each density function. Each computation was performed on the big grid described in Table \ref{tab:grids}, for different values of the distance-size ratio $D$. If the difference is less than 0.1\%, we consider that the model has achieved an acceptable accuracy for the corresponding value of $D$.
}
\label{fig:D-exp-power-thin}
\end{figure}

@@

\noindent where $\rho_1$ and $\rho_2$ are constant densities, $R_1$ and $R_2$ are the inner and outer radii of the shell, respectively, and $R_c$ is the the radius on which the density change takes place.

The analytical solution for this kind of density is a linear combination of two homogeneous shells with densities $\rho_1$ and $\rho_2$:

@@

Firstly, lets assume a symmetric density function, i.e. $R_c=(R_1 + R_2)/2$ is the centre radius of the shell, and perform the test for a thin shell of 1km of thickness without applying the density-based discretization algorithm.
The results can be seen in Figure \ref{fig:D-discont-sym}.
Once again, we obtain good accuracy for values of $D$ equal to 1, 2 and 8 for the computation of the potential, the gradient components and the Marussi tensor components, respectively.

@@

As we did with the high varying exponential densities, we can conclude that for arbitrary discontinuous functions a density-based discretization algorithm is also needed in order to guarantee a good performance.
